% CAIS 2026 Workshop Extended Abstract
% ACM Conference on AI Safety, May 27-29, San Jose
% Draft v7.2: April 10, 2026
% V7.2: Fix Cohen's d, run counts, capitalize Bloom.
% V7.1: Add positive coupling sign-flip confirmation (320 additional runs).
% V7: Integrate Claude Chorus LLM experiment, remove bridge (for evolving-strategies),
%   add contravariant functor explanation, explicit agent definitions.
\documentclass[10pt,twocolumn]{article}

\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage[T1]{fontenc}
\usepackage[hyphens]{url}
\usepackage[expansion=false]{microtype}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{enumitem}
\setlist{nosep,leftmargin=*}

\newtheorem{definition}{Definition}

\graphicspath{{figures/}}

\title{What Your Agent Topology Is Actually Doing:\\
Controlled Experiments on Communication Structure\\
in Multi-Agent Systems}

\author{
  Robin Langer \and Lyra Vega
}

\date{}

\begin{document}
\maketitle

\begin{abstract}
Multi-agent AI systems choose communication topologies---star, chain,
ring, fully connected---with no controlled evidence for how topology
affects behavior. We provide three lines of evidence. First, using
island-model genetic algorithms as a model system (240~runs), we show
topology independently explains 17\% of behavioral diversity variance
($p < 10^{-7}$), amplified $14\times$ on hard problems. Second, we
validate the transfer to real agents: 3,200 LLM~API calls across
8~coding tasks show that topology determines solution diversity in
GPT-4o-mini agents. Under negative coupling (``be different''), the
diversity ordering \emph{inverts} from GAs (Cohen's $d$ from $-14.0$ to $-1.9$ across six tasks);
under positive coupling (``build on and improve''), it reverts to
GA-like ordering ($\eta^2 = 0.291$, $p < 10^{-19}$). Changing one
word in the prompt flips which topology produces the most diverse
output. The sign depends on which adjoint of the topology the coupling
mechanism instantiates---left Kan (homogenize) vs.\ right Kan
(diversify). At least nine deployed systems chose topologies without
controlled comparisons. Topology is a first-class safety variable;
current systems are tuning it blindly.
\end{abstract}

%% ===================================================================
\section{Introduction}
\label{sec:intro}
%% ===================================================================

Every multi-agent system embeds a communication topology---round-robin
group chat (AutoGen), DAG pipelines (LangGraph), manager hubs
(CrewAI)---yet these choices are made without controlled evidence.
Dochkina~\citep{dochkina2026} ran 25,000 tasks across 256 agents,
finding a 44\% quality spread ($d = 1.86$) between topologies.
Hierarchical deployments fail 36\% of the time; swarms fail 68\%.
Topology is doing something---but \emph{what, exactly?}

We provide the first controlled answer, bridging from a model system
(genetic algorithms) to real LLM agents. Our contributions:
(1)~topology independently determines behavioral diversity in both GAs
and LLM agents; (2)~the effect \emph{inverts} between GAs and
agents, explained by which adjoint of the topology the coupling
mechanism instantiates; (3)~at least nine deployed systems chose
topologies without controlled comparisons.

%% ===================================================================
\section{Island Models as Agent Systems}
\label{sec:analogy}
%% ===================================================================

An island-model GA partitions a population across sub-populations
(\emph{islands}) that evolve independently, exchanging individuals
periodically via \emph{migration} channels. The mapping to agent
orchestration is direct:

\begin{center}
\small
\begin{tabular}{@{}p{3.2cm}p{3.8cm}@{}}
\toprule
\textbf{GA concept} & \textbf{Agent equivalent} \\
\midrule
Island & Agent (local state, computation) \\
Migration channel & Message channel \\
Population diversity & Solution diversity \\
Fitness function & Task performance metric \\
Migration topology & Orchestration graph \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The analogy is structural, not metaphorical: both systems have nodes
with local state exchanging information along directed channels.
The agent-side quantities are measured directly:
\textbf{agent diversity} = mean pairwise cosine distance on sentence
embeddings (all-MiniLM-L6-v2) of agent outputs;
\textbf{agent fitness} = task performance (test pass rate,
correctness). GAs provide cheap experimental control (hundreds of
trials); the LLM experiment (Section~\ref{sec:chorus}) validates
transfer to real agents.

%% ===================================================================
\section{Definitions}
\label{sec:definitions}
%% ===================================================================

\begin{definition}[Communication topology]
A \emph{communication topology} for $n$ agents is a directed graph
$D = (V, A)$ with $|V| = n$, where each arc $(u, v) \in A$ is a
permitted channel from agent $u$ to agent $v$.
\end{definition}

\begin{definition}[Cycle rank]
The \emph{cycle rank} $\beta_1(G) = |E| - |V| + 1$ (for connected
$G$) counts independent cycles---feedback loops in the communication
structure.
\end{definition}

\begin{definition}[Directed cycle count]
$\kappa(D)$ counts distinct simple directed cycles in
$D$~\citep{johnson1975finding}. Unlike $\beta_1$, this depends on arc
\emph{arrangement}, not just edge count. A DAG has $\kappa = 0$.
\end{definition}

\paragraph{NK landscapes}~\citep{kauffman1993origins} are tunably
rugged fitness functions over binary strings of length $N$, where each
bit's contribution depends on $K$ other bits. $K{=}0$ gives a smooth
landscape; increasing $K$ adds local optima.

%% ===================================================================
\section{Experimental Evidence}
\label{sec:experiments}
%% ===================================================================

We present three experiments spanning GA model systems and LLM agents,
totaling 490 GA runs and 3,200 LLM~API calls.

\subsection{Experiment 1: Topology affects diversity at constant density}

Comparing a chain to a complete graph conflates density and cycle
structure: $\beta_1 = m - n + 1$~\citep{hatcher2002algebraic} means
cycle count and edge count are perfectly correlated for undirected
graphs. We escape via directed cycle count $\kappa$, which varies
independently of density.

\paragraph{Design.}
8 directed graphs, all $n{=}8$, 16 arcs ($\delta{=}2.0$), varying
only in $\kappa \in \{0, 0, 3, 10, 14, 20, 29, 47\}$.
Island-model GA: 50 individuals/island, 500 generations, 30 seeds
per topology, 240 runs total.

\paragraph{Results.}
$F(7, 232) = 6.90$, $p = 1.85 \times 10^{-7}$, $\eta^2 = 0.17$.
More directed cycles \emph{accelerate consensus} ($r = -0.68$):
topologies with more feedback loops cause agents to converge faster,
reducing capacity to maintain alternative strategies.

\subsection{Experiment 2: Hard problems amplify the effect $14\times$}

On NK landscapes with tunable ruggedness (250 runs across 4 values of
$K$, 2--3 topologies, 10--20 seeds), topology's effect scales with
difficulty:

\begin{center}
\small
\begin{tabular}{@{}ccc@{}}
\toprule
$K$ & $\eta^2$ (diversity) & Interpretation \\
\midrule
0 (trivial) & 0.05 & Topology barely matters \\
2 (moderate) & 0.20 & Moderate effect \\
4 (hard) & 0.45 & Large effect \\
6 (very hard) & 0.69 & Topology dominates \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Simple tasks are structurally smooth---topology barely matters.
Complex tasks (multiple valid strategies, ambiguous evaluation) are
rugged---topology dominates. Real-world deployments overwhelmingly
involve the latter.

\subsection{Experiment 3: Topology inverts diversity in LLM agents}
\label{sec:chorus}

The GA experiments establish that topology controls diversity.
But does this transfer to LLM agents? We test directly.

\paragraph{Design.}
5 GPT-4o-mini agents, 8 coding tasks (shortest path, kth largest,
rate limiter, JSON parser, LRU cache, regex matcher, expression
evaluator, Bloom filter), 4 topologies (none, ring, star, fully
connected), 10 seeds per condition. 1,600 API calls total.
Coupling is \emph{negative}: each agent is instructed to use a
\emph{different} algorithm from those visible in its context. Diversity
is measured as mean pairwise cosine distance on sentence embeddings
(all-MiniLM-L6-v2) of agent outputs.

\paragraph{Results.}
The diversity ordering \textbf{inverts} from the GA result:

\begin{center}
\small
\begin{tabular}{@{}llp{3.6cm}@{}}
\toprule
\textbf{System} & \textbf{Diversity ordering} & \textbf{Statistic} \\
\midrule
GA & none $>$ ring $>$ FC & $W = 1.0$ (6 domains) \\
LLM & FC $>$ ring $\approx$ star $>$ none & $W = 0.91$ (6/8 tasks) \\
\bottomrule
\end{tabular}
\end{center}

\noindent
Six of eight tasks show the inverted ordering cleanly (Kruskal-Wallis
$p < 10^{-6}$; Cohen's $d$ ranges from $-14.8$ to $-1.9$). Two
exceptions are informative: Bloom filter ($d \approx 0$, unimodal
task---no algorithmic variety for topology to act on) and expression
evaluator ($d = +0.4$, the model's prior is already spread).

\paragraph{Why the sign flips.}
Under ``none'' topology, all 5 LLM agents independently generate
the \emph{same} algorithm (e.g., quickselect for kth\_largest).
LLMs start from a \textbf{peaked prior} (low entropy); GAs start
from \textbf{random initialization} (high entropy). Isolation
preserves initial entropy: high for GAs (diverse), low for LLMs
(homogeneous). The inversion follows from which \emph{adjoint} of the
topology the coupling mechanism instantiates:

\begin{itemize}
  \item \textbf{GA migration} = \textbf{left Kan extension}
    (colimit-like): merges populations. More connectivity $\to$ less
    diversity.
  \item \textbf{LLM ``be different'' context} = \textbf{right Kan
    extension} (limit-like): accumulates exclusion constraints. More
    connectivity $\to$ agents forced into distributional tails $\to$
    \emph{more} diversity.
\end{itemize}

\noindent
The topology is the invariant; the sign depends on the adjoint.
$\mathrm{Lan}_F \dashv \mathrm{Ran}_F$ reverses ordering on any poset,
including the diversity ordering on topologies.

\paragraph{Positive coupling confirms the prediction.}
We repeated the experiment with \emph{positive} coupling: each agent
is instructed to ``build on and improve'' neighbors' solutions instead
of ``be different.'' 1,600 additional API calls (total: 3,200). Results:
the diversity ordering \textbf{reverts to GA-like}: none $>$ star
$\approx$ ring $>$ FC (Kruskal-Wallis $H = 94.83$, $p = 2.01 \times
10^{-20}$; Cohen's $d = 1.08$, none vs.\ FC; $\eta^2 = 0.291$).
Crucially, the \emph{none} topology---where agents have no
neighbors---shows identical diversity across conditions (negative:
0.090, positive: 0.095, $p = 0.96$), confirming that coupling sign,
not some other experimental variable, drives the reversal. Connected
topologies collapse to near-zero diversity (0.025--0.028) under
positive coupling, as agents converge on a shared best solution. One
word change in the prompt flips the diversity ordering---precisely as
the $\mathrm{Lan} \dashv \mathrm{Ran}$ framework predicts.

%% ===================================================================
\section{Implications for Agent Orchestration}
\label{sec:implications}
%% ===================================================================

\paragraph{The evidence gap.}
At least nine recent agent systems---spanning ICML, AAAI, WWW, and
NeurIPS---chose topologies without density-controlled comparisons
(Table~\ref{tab:systems}).

\begin{table}[t]
\centering
\caption{Agent systems whose topology choices lack controlled evidence.}
\label{tab:systems}
\footnotesize
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}lp{3.5cm}@{}}
\toprule
\textbf{System} & \textbf{Issue} \\
\midrule
Graph-GRPO~\citep{graphgrpo2026} & Complete = worst; also max $\beta_1$ \\
OFA-MAS~\citep{ofamas2026} & Prunes density \& cycles together \\
G-Designer~\citep{gdesigner2025} & No density-controlled baseline \\
GoAgent~\citep{goagent2026} & DAG vs.\ dense alternatives \\
ARG-Designer~\citep{argdesigner2026} & DAG space only \\
AgentConductor~\citep{agentconductor2026} & Cycles never tested \\
DyTopo~\citep{dytopo2026} & DAG enforced throughout \\
HERA~\citep{hera2026} & Cycles tracked, not controlled \\
Dochkina~\citep{dochkina2026} & Density \& cycles confounded \\
\bottomrule
\end{tabular}}
\end{table}

\paragraph{Safety risks.}
Our LLM experiment sharpens three concerns:
\emph{(1)~The sign matters.} Changing one word in the coupling
instruction flips which topology produces the most diverse output
($\eta^2 = 0.291$, $p < 10^{-19}$). Predictions based on GA
intuition alone will have the wrong sign for diversifying agents.
\emph{(2)~Premature consensus.} Isolated LLM agents converge to
identical solutions (Cohen's $d$ from $-14.0$ to $-1.9$ vs.\ connected agents across six tasks).
``Ensemble diversity'' claims based on independent agents may be
illusory when models share a peaked prior.
\emph{(3)~Error amplification.} Moran~\citep{moran2025error} reports
$17\times$ error amplification in unstructured communication. The
standard fix (impose DAG structure) simultaneously removes cycles
\emph{and} density---we cannot know which helps.

\paragraph{Recommendations.}
\begin{enumerate}
  \item \textbf{Treat topology as a first-class variable.}
    It controls the diversity--convergence tradeoff---in both GAs and
    LLM agents, empirically confirmed.
  \item \textbf{Know your coupling sign.} One word change in the
    coupling instruction flips the diversity ordering empirically.
    Building on neighbors' outputs (left Kan) homogenizes;
    differentiating from them (right Kan) diversifies.
  \item \textbf{Match topology to task complexity.} Simple tasks
    tolerate any topology; hard tasks demand deliberate choice
    ($\eta^2$ scales $14\times$ from $K{=}0$ to $K{=}6$).
  \item \textbf{Monitor diversity, not just accuracy.} Topology
    effects manifest in strategy diversity. Accuracy alone masks
    topology-induced failures.
\end{enumerate}

%% ===================================================================
\section{Related Work}
\label{sec:related}
%% ===================================================================

Yang et al.~\citep{yang2025topology} identify topology as the
underexplored MAS dimension; we provide the controlled experiments
they call for. Dochkina~\citep{dochkina2026} shows massive effects
(25K tasks, $d = 1.86$) without isolating the mechanism.
HERA~\citep{hera2026} tracks cycle count during evolution but does not
control density. In evolutionary computation, the topology--diversity
link is established~\citep{tomassini2005spatially,skolicki2005influence}
without density control. Capucci and
Myers~\citep{capucci2026feedback} formalize the cost: each cycle
requires an independent convergence guarantee.

%% ===================================================================
\section{Conclusion}
\label{sec:conclusion}
%% ===================================================================

Communication topology is not neutral infrastructure---it is a design
variable controlling how multi-agent systems balance exploration
against convergence. In GA model systems, topology independently
explains 17\% of diversity variance on easy problems, rising to 69\%
on hard ones. In LLM agents, the effect is equally strong but its \emph{sign}
depends on coupling: negative coupling (``be different'') inverts the
GA ordering ($d$ from $-14.0$ to $-1.9$); positive coupling (``build on'') restores
it ($\eta^2 = 0.291$). The sign is determined by which adjoint the
coupling instantiates---left Kan (homogenize) vs.\ right Kan
(diversify)---confirmed by the null control: isolated agents show
identical diversity regardless of coupling instruction ($p = 0.96$). At least nine
deployed systems chose topologies based on comparisons that
simultaneously varied the property they claimed to test. For safe
multi-agent deployment, topology must be tuned deliberately, its
coupling sign understood, and its effects monitored through diversity
metrics---not assumed away.

\smallskip
\noindent\textbf{Acknowledgments.} [Withheld for review.]

%% ===================================================================
%% References
%% ===================================================================

\bibliographystyle{plainnat}
\begin{thebibliography}{20}

\bibitem[Capucci and Myers(2026)]{capucci2026feedback}
M.~Capucci and D.~J. Myers.
\newblock Feedback in categorical multi-agent systems.
\newblock \emph{arXiv:2604.03017}, 2026.

\bibitem[Dochkina(2026)]{dochkina2026}
V.~Dochkina.
\newblock Drop the hierarchy and roles: How self-organizing {LLM} agents
  outperform designed structures.
\newblock \emph{arXiv:2603.28990}, 2026.

\bibitem[Chen et~al.(2026)]{goagent2026}
H.~Chen et~al.
\newblock GoAgent: Group-of-agents communication topology generation for
  {LLM}-based multi-agent systems.
\newblock \emph{arXiv:2603.19677}, 2026.

\bibitem[Wang et~al.(2026)]{agentconductor2026}
S.~Wang et~al.
\newblock AgentConductor: Topology evolution for multi-agent competition-level
  code generation.
\newblock \emph{arXiv:2602.17100}, 2026.

\bibitem[Cang et~al.(2026)]{graphgrpo2026}
Y.~Cang et~al.
\newblock Graph-GRPO: Stabilizing multi-agent topology learning via group
  relative policy optimization.
\newblock \emph{arXiv:2603.02701}, 2026.

\bibitem[Li et~al.(2026)]{ofamas2026}
S.~Li et~al.
\newblock OFA-MAS: One-for-all multi-agent system topology design based on
  mixture-of-experts graph generative models.
\newblock In \emph{Proc.\ WWW}, 2026.

\bibitem[Zhang et~al.(2025)]{gdesigner2025}
G.~Zhang et~al.
\newblock G-Designer: Architecting multi-agent communication topologies via
  graph neural networks.
\newblock In \emph{Proc.\ ICML}, 2025.

\bibitem[Li et~al.(2026b)]{argdesigner2026}
S.~Li et~al.
\newblock Assemble your crew: Automatic multi-agent communication topology
  design via autoregressive graph generation.
\newblock In \emph{Proc.\ AAAI}, 2026.

\bibitem[Hatcher(2002)]{hatcher2002algebraic}
A.~Hatcher.
\newblock \emph{Algebraic Topology}.
\newblock Cambridge University Press, 2002.

\bibitem[Li and Ramakrishnan(2026)]{hera2026}
S.~Li and N.~Ramakrishnan.
\newblock Experience as a compass: Multi-agent {RAG} with evolving
  orchestration and agent prompts.
\newblock \emph{arXiv:2604.00901}, 2026.

\bibitem[Johnson(1975)]{johnson1975finding}
D.~B. Johnson.
\newblock Finding all the elementary circuits of a directed graph.
\newblock \emph{SIAM J.\ Comput.}, 4(1):77--84, 1975.

\bibitem[Kauffman(1993)]{kauffman1993origins}
S.~A. Kauffman.
\newblock \emph{The Origins of Order}.
\newblock Oxford University Press, 1993.

\bibitem[Moran(2025)]{moran2025error}
K.~Moran.
\newblock The secret is topology: Why unstructured multi-agent communication
  amplifies errors.
\newblock \emph{Towards Data Science}, 2025.

\bibitem[Skolicki and De~Jong(2005)]{skolicki2005influence}
Z.~Skolicki and K.~A. De~Jong.
\newblock The influence of migration sizes and intervals on island models.
\newblock In \emph{GECCO}, pages 1295--1302, 2005.

\bibitem[Tomassini(2005)]{tomassini2005spatially}
M.~Tomassini.
\newblock \emph{Spatially Structured Evolutionary Algorithms}.
\newblock Springer, 2005.

\bibitem[Yang et~al.(2025)]{yang2025topology}
J.~Yang et~al.
\newblock Topological structure learning should be a research priority for
  {LLM}-based multi-agent systems.
\newblock \emph{arXiv:2505.22467}, 2025.

\bibitem[Lu et~al.(2026)]{dytopo2026}
Y.~Lu et~al.
\newblock DyTopo: Dynamic topology routing for multi-agent reasoning via
  semantic matching.
\newblock \emph{arXiv:2602.06039}, 2026.

\end{thebibliography}

\end{document}
