\documentclass[10pt]{article}

\usepackage[margin=0.85in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage{enumitem}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

\setlength{\parskip}{0.3em}

\title{\vspace{-1.5em}Genetic Algorithms as Kleisli Morphisms:\\
  Compositional Invariants Across Evolutionary Domains\vspace{-0.5em}}

\author{
  Lyra \and Claudius \and Robin Langer \and Nick Meinhold
}

\date{}

\begin{document}

\maketitle
\vspace{-1em}

\noindent
\textbf{Keywords:} Kleisli categories, genetic algorithms, lax monoidal functors, compositional invariants

\medskip

Lax monoidal functors do not preserve composition strictly. Ghani et al.~\cite{ghani2024} demonstrated this for open games: the approximation functor $T_\varepsilon$ is lax monoidal, so approximate Nash equilibria fail to compose like exact ones. We show the same phenomenon arises in evolutionary computation: the island model functor on genetic algorithms is lax monoidal whenever migration occurs. This is not an analogy --- both are instances of the same abstract failure mode, in different categories.

Hanks et al.~\cite{hanks2024} showed that gradient descent is a hypergraph functor from optimization problems to dynamical systems. Their framework requires differentiability. We provide the categorical foundation for the \emph{gradient-free} regime: population-based evolutionary search, where the relevant structure is a monad carrying randomness, configuration, and logging.

\textbf{Framework.}
A genetic algorithm evaluates fitness, selects survivors, recombines, and mutates. These operators require random number generation, read-only parameters, and write-only logging --- three effects packaged into a monad $M = \mathsf{ReaderT}\;\mathit{Config}\;(\mathsf{WriterT}\;\mathit{Log}\;(\mathsf{State}\;\mathit{StdGen}))$. A \emph{genetic operator} is a Kleisli arrow $\mathit{op} \colon [a] \to M\,[b]$ on populations. One generation is the Kleisli composite
\[
  \mathit{step} = \mathit{evaluate} \ggg \mathit{select} \ggg \mathit{crossover} \ggg \mathit{mutate} \;\colon\; [a] \to M\,[a]
\]
and an $n$-generation run is $\mathit{step}^n$. This instantiates Miranda's abstract Kleisli structures~\cite{miranda2024} concretely: the Kleisli category $\mathcal{K}\!\ell(M)$ with population types as objects and GA operators as morphisms. We identify three composition levels --- operators into pipelines, pipelines into strategies, strategies into multi-population systems --- each forming a category with the same Kleisli composition.

\textbf{Result 1: Diversity Fingerprints.}
Different strategy compositions produce qualitatively distinct diversity trajectories \emph{stable across genome types and fitness landscapes}. On three domains (real-valued weight vectors, binary mazes, expression trees), we identify four fingerprints: monotonic decline (generational), spike--crash--rebound (hourglass), stable maintenance (island), and spike--collapse (adaptive). These are determined by the \emph{composition pattern}, not by the operators or genome type. We conjecture, supported by consistent experimental patterns across all three domains, that diversity fingerprints are functorial invariants of strategy composition, analogous to Myers' behavioral invariants of composed dynamical systems~\cite{myers2022}.

\textbf{Result 2: The Strict/Lax Dichotomy.}
The island model lifts a single-population strategy to a multi-population strategy with periodic migration, defining a functor $I \colon \mathbf{Strat} \to \mathbf{Strat}$.

\begin{theorem}[Strict/Lax Dichotomy]
For strategies $S_1, S_2$ and island functor $I_{\mu}$ with migration rate $\mu$:
\begin{enumerate}[nosep,leftmargin=1.5em]
  \item If $\mu = 0$: $I(S_1 \ggg S_2) = I(S_1) \ggg I(S_2)$ \emph{(strict monoidal)}.
  \item If $\mu > 0$: the laxator $\lambda \colon I(S_1) \ggg I(S_2) \Rightarrow I(S_1 \ggg S_2)$ has magnitude $D^*$ \emph{(population divergence at composition boundaries)} that is independent of boundary position and migration frequency.
\end{enumerate}
\end{theorem}

\noindent
We validate this across all three domains (checkers, mazes, symbolic regression). The laxity arises because resetting the migration schedule at the composition boundary perturbs a chaotic dynamical system; the perturbation magnitude is determined by the spectral gap of the migration topology's Markov chain, not by where or how often migration occurs. Both this result and Ghani et al.'s~\cite{ghani2024} instantiate a general pattern: \emph{relaxation of exactness conditions produces laxity whose magnitude is independent of the relaxation parameters}.

\textbf{Result 3: Three-Level Composition.}
The operator $\to$ pipeline $\to$ strategy tower is categorical: each level forms a category, inclusions are functors, strategy combinators are endofunctors on $\mathbf{Strat}$, and migration is a natural transformation between island configurations. This enables reasoning by induction on composition depth: strategy equivalence reduces to operator-level equivalence, making correctness proofs for large evolutionary systems tractable.

\textbf{Positioning.}
Evolutionary computation has never appeared at ACT. We connect it to three active threads: (1)~Ghani's compositional game theory~\cite{ghani2024} --- the Dichotomy Theorem is the evolutionary instance of lax composition of approximate equilibria; (2)~Capucci et al.'s categorical cybernetics~\cite{capucci2022} --- GAs are bidirectional processes (forward: genetic operators; backward: fitness-based selection) that may admit an optic description, complementing Hedges and Sakamoto's treatment of RL~\cite{hedges2024}; (3)~Hanks et al.'s optimization-as-functor~\cite{hanks2024} --- we cover the gradient-free regime. The framework is implemented in Haskell (12 modules, 75 tests, ${\sim}5500$ lines). A full manuscript is in preparation.

\begin{thebibliography}{9}
\small
\setlength{\itemsep}{0pt}

\bibitem{ghani2024}
N.~Ghani, J.~Hedges, V.~Winschel, P.~Zahn. A category theoretic approach to approximate game theory. \emph{EPTCS} 429, 2024. arXiv:2509.20932.

\bibitem{hanks2024}
T.~Hanks, J.~Fairbanks, M.~Klawonn. Generalized gradient descent is a hypergraph functor. \emph{EPTCS} 429, pp.~217--233, 2025. doi:10.4204/EPTCS.429.12.

\bibitem{miranda2024}
A.~Miranda. Abstract Kleisli structures on 2-categories. \emph{EPTCS} 429, pp.~29--48, 2025. doi:10.4204/EPTCS.429.2.

\bibitem{capucci2022}
M.~Capucci, B.~Gavranovi\'c, J.~Hedges, E.~F.~Rischel. Towards foundations of categorical cybernetics. \emph{EPTCS} 372, pp.~235--248, 2022. arXiv:2105.06332.

\bibitem{hedges2024}
J.~Hedges, R.~Rodriguez Sakamoto. Reinforcement learning in categorical cybernetics. \emph{EPTCS} 429, pp.~270--286, 2025. arXiv:2404.02688.

\bibitem{myers2022}
D.~J. Myers. Double categories of open dynamical systems. \emph{Proc.\ ACT} 2020. arXiv:2005.05956.

\end{thebibliography}

\end{document}
