\documentclass[sigconf,anonymous]{acmart}

% Remove default ACM copyright for workshop submission draft
\setcopyright{none}
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}

% amsmath, amssymb, amsthm, booktabs are loaded by acmart
\usepackage{tikz-cd}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\begin{document}

\title{Composition Determines Behavior: Diversity Fingerprints and the Strict/Lax Dichotomy in Genetic Algorithms}

\author{Claudius Turing}
\affiliation{%
  \institution{Open University}
  \city{Helsinki}
  \country{Finland}}

\author{Lyra Vega}
\email{lyraclaude20@gmail.com}
\affiliation{%
  \institution{Open University}
  \city{Helsinki}
  \country{Finland}}

\begin{abstract}
How you compose GA operators matters more than which operators you choose. We demonstrate this empirically across four domains---checkers (real-valued), mazes (binary), symbolic regression (tree-structured), and OneMax (binary)---by formalizing GA pipelines as morphisms in a Kleisli category, where the monad carries randomness, configuration, and logging. In this framework, a full evolutionary run is a single composite arrow, and strategies (generational, island, hourglass, adaptive) are higher-order operations on these arrows. We present three results. First, different strategy compositions produce qualitatively distinct diversity trajectories---\emph{diversity fingerprints}---that are stable across genome types and fitness landscapes; a topology sweep across five additional domains---including a co-evolutionary domain with no fixed fitness landscape---confirms identical ordering with topology explaining $23.9\times$ more variance than domain ($p_{\mathrm{domain}} = 0.945$). Second, the island strategy functor preserves sequential composition if and only if no migration occurs: one migrant changes everything (the \emph{Strict/Lax Dichotomy}). Across 30 independent seeds, strict and lax compositions diverge with a large effect size (Cohen's $d = 4.34$, $p < 4 \times 10^{-11}$) while converging to identical fitness optima. Third, operators compose into pipelines, pipelines into strategies, and strategies into multi-population systems, each level sharing the same algebraic structure. The composition pattern, not the individual operators, determines the evolutionary trajectory.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003809.10010031</concept_id>
<concept_desc>Theory of computation~Categorical semantics</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010147.10010178.10010187</concept_id>
<concept_desc>Computing methodologies~Genetic algorithms</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Categorical semantics}
\ccsdesc[500]{Computing methodologies~Genetic algorithms}

\keywords{genetic algorithms, composition, diversity, population dynamics, category theory}

\maketitle

\section{Introduction}

How you compose GA operators matters more than which operators you choose. This claim runs against decades of evolutionary computation research focused on designing better selection schemes~\cite{blickle1996comparison}, better crossover operators~\cite{eshelman1993crossover}, and better mutation strategies~\cite{back1996evolutionary}. Individual operators matter, of course. But when we compose the same operators into different high-level strategies---generational, island, hourglass, adaptive---the \emph{composition pattern} determines the population's diversity trajectory. Swap out tournament selection for roulette: the trajectory barely changes. Switch from generational to island composition: the trajectory transforms completely.

GA practitioners already compose operators routinely. Every generational GA runs select $\to$ crossover $\to$ mutation $\to$ replacement as a pipeline. Island models compose multiple such pipelines with periodic migration. Adaptive strategies compose phases with runtime switching. But this composition is typically buried inside imperative loops, invisible to formal analysis. We cannot ask ``does this composition preserve a property of its parts?'' because the composition has no formal representation.

We make composition explicit by formalizing GA operators as morphisms in a \emph{Kleisli category}---a standard construction from programming language theory~\cite{moggi1991notions} that gives compositionality for effectful computations. Every GA operator reads configuration, logs metrics, and uses randomness. The Kleisli category lets us compose these effectful operators as if they were pure functions, while a monad threads the effects automatically. A full evolutionary run becomes a single composite arrow. Strategies become higher-order operations on arrows.

This formalization is not merely notational. It yields three concrete results:

\begin{enumerate}[nosep]
  \item \textbf{Diversity fingerprints.} Different strategy compositions produce qualitatively distinct diversity trajectories: monotonic decline (generational), spike-crash-rebound (hourglass), steady maintenance (island), and spike-then-collapse (adaptive). These fingerprints are compositionally determined and stable across genome types and fitness landscapes.

  \item \textbf{The Strict/Lax Dichotomy.} The island strategy preserves sequential composition strictly if and only if no migration occurs. Any nonzero migration---even one individual per thousand generations---breaks strict preservation. The magnitude of divergence is independent of migration frequency, a consequence of chaotic amplification at composition boundaries.

  \item \textbf{Three-level composition tower.} Operators compose into pipelines (one generation), pipelines compose into strategies (a full run), and strategies compose into multi-population systems. Each level uses the same Kleisli composition, giving a recursive design algebra for building complex GAs from simple parts.
\end{enumerate}

We demonstrate the framework across three domains chosen for maximum dissimilarity: checkers evaluation (real-valued weight vectors), maze topology (binary genomes), and symbolic regression (expression trees). The categorical structure is invariant across all three---only the genome types and domain-specific operators change.

The paper proceeds as follows. Section~2 introduces the compositional framework: the evolution monad, operators as morphisms, and the three composition levels. Section~3 describes the experimental setup. Section~4 presents results on diversity fingerprints, cross-domain stability, and the Strict/Lax Dichotomy. Section~5 surveys related work. Section~6 discusses practical implications and future directions.

\section{Framework: Compositional Genetic Algorithms}

The central idea is simple: treat GA operators as composable arrows in a category where the composition mechanism handles side effects automatically. We build this in three steps: define the effect context (the monad), type the operators (the morphisms), and identify three levels at which composition occurs.

\subsection{The Evolution Monad}

Every GA operator performs three kinds of side effects. Selection, crossover, and mutation all draw random numbers. Every operator reads configuration parameters (tournament size, mutation rate, population size). And every generation produces statistics that accumulate into a run log. In imperative code, these effects are threaded manually---a random seed passed as \texttt{\&mut rng}, a config struct accessed everywhere, a log vector appended to each iteration.

We bundle all three effects into a single monad stack $M$:

\begin{verbatim}
  type EvoM = ReaderT GAConfig
               (WriterT GALog
                 (State StdGen))
\end{verbatim}

\noindent In plain English: \texttt{EvoM} is a computation that (1)~reads a \texttt{GAConfig} without modifying it, (2)~appends to a \texttt{GALog} that accumulates over the run, and (3)~threads a random number generator state. Any function running inside \texttt{EvoM} has access to all three effects without explicitly passing them as arguments.

A \emph{Kleisli arrow} for this monad has the form $a \to M\,b$: a function that takes an input of type $a$ and produces an output of type $b$ wrapped in the effect context $M$. Kleisli composition, written $\mathbin{>\!\!=\!\!>}$, chains two such arrows: apply the first, extract the intermediate result from $M$, feed it to the second, while $M$'s internal machinery threads the random seed, propagates the configuration, and accumulates the log. The individual operators never see this plumbing.

This is the Kleisli category of $M$---a standard construction from category theory~\cite{maclane1971categories, moggi1991notions} that gives us compositionality for free. The monad laws (associativity and identity) guarantee that our compositions are well-behaved: re-parenthesizing a pipeline does not change its behavior, and inserting a do-nothing operator has no effect.

\subsection{Operators as Morphisms}

With the monad in place, we type the core GA operators as Kleisli arrows operating on populations. Let \texttt{Pop\,a} denote a population of individuals with genome type~$a$:

\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Operator} & \textbf{Type} \\
\midrule
Selection   & \texttt{Pop\,a $\to$ $M$\,(Pop\,a)} \\
Crossover   & \texttt{Pop\,a $\to$ $M$\,(Pop\,a)} \\
Mutation    & \texttt{Pop\,a $\to$ $M$\,(Pop\,a)} \\
Replacement & \texttt{Pop\,a $\to$ $M$\,(Pop\,a)} \\
\bottomrule
\end{tabular}
\end{center}

\noindent Every operator has the uniform Kleisli arrow type \texttt{Pop\,a $\to$ $M$\,(Pop\,a)}: one population in, one population out, with effects bundled in~$M$. Selection reads tournament size from the config and draws random competitors. Crossover reads the crossover rate and generates random split points. Mutation reads the mutation rate and perturbs random individuals. Replacement merges offspring with the parent population; it accesses the parent population through the \texttt{Reader} component of the monad stack rather than taking it as a second argument, preserving the uniform Kleisli signature. All four operators log diversity statistics.

A \emph{pipeline} is the Kleisli composition of these operators into one generation step:

\begin{verbatim}
  pipeline = selection >=> crossover
                       >=> mutation
                       >=> replacement
\end{verbatim}

\noindent This composite has the same type as its parts: \texttt{Pop\,a $\to$ $M$\,(Pop\,a)}. It is a single arrow that, given a population and a random seed, produces a new population and a log of what happened. The pipeline is itself an operator---it can be composed further.

Consider a concrete example. The generational strategy applies the pipeline $n$~times:

\begin{verbatim}
  generational n = pipeline >=> pipeline
                             >=> ... (n times)
\end{verbatim}

\noindent Each application threads the monad: the random seed advances, the log grows, the config persists. The result is one composite arrow: $n$~generations of evolution, packaged as a single composable unit.

The genome type $a$ is a parameter. For checkers, $a = \texttt{[Double]}$ (weight vectors). For mazes, $a = \texttt{[Bool]}$ (binary genomes). For symbolic regression, $a = \texttt{Expr}$ (expression trees). The pipeline shape is identical across all three---only the concrete implementations of \texttt{mutation} and \texttt{crossover} change (Gaussian perturbation vs.\ bit flip vs.\ subtree swap). This is precisely what we mean by ``composition determines behavior'': the same compositional structure produces the same diversity dynamics regardless of the underlying genome representation.

\subsection{Composition Levels}

The framework admits three levels of composition, each using the same Kleisli mechanism:

\paragraph{Level 1: Operators $\to$ Pipelines.}
Individual operators (selection, crossover, mutation, replacement) compose into a pipeline representing one generation. This is the level described above. The pipeline is a single Kleisli arrow.

\paragraph{Level 2: Pipelines $\to$ Strategies.}
Pipelines compose into strategies that define the overall structure of an evolutionary run. Four strategies illustrate the design space:

\begin{itemize}[nosep]
  \item \textbf{Generational}: Apply the pipeline $n$ times sequentially. Pure iteration.
  \item \textbf{Hourglass}: Compose three phases---explore (high mutation), converge (strong selection), diversify (high mutation again)---each a pipeline with different parameters.
  \item \textbf{Island}: Run $k$ independent pipelines in parallel, connected by periodic migration. The island construction is a functor: it transforms a single-population pipeline into a multi-population pipeline while preserving (or laxly preserving) composition.
  \item \textbf{Adaptive}: Compose an exploration phase with a focused phase, switching at runtime when a plateau is detected. A conditional composition with a predicate.
\end{itemize}

\noindent Each strategy is itself a Kleisli arrow \texttt{Pop\,a $\to$ $M$\,(Pop\,a)}. It can be composed with other strategies, nested inside an island model, or used as a building block for more complex designs.

\paragraph{Level 3: Strategies $\to$ Multi-Population Systems.}
Strategies compose into multi-population architectures. An island model containing four generational GAs with migration is a composition at this level. So is a heterogeneous system where different islands run different strategies (one generational, one adaptive) and exchange individuals.

The recursive structure is the key insight. At every level, the composite has the same type as its parts: a Kleisli arrow on populations. A pipeline of five operators is itself an operator. A strategy of three phases is itself a strategy. An island of four strategies is itself a strategy. This closure under composition gives us a \emph{design algebra} for genetic algorithms: complex systems are built by composing simple parts, and the properties of the composite can be analyzed in terms of how its parts are composed---not what those parts do internally.

This algebraic view directly enables our main results. Diversity fingerprints (Section~4.1) are properties of composition patterns, not individual operators. The Strict/Lax Dichotomy (Section~4.3) is a statement about when a specific composition---the island functor---preserves the structure of sequential composition. Neither result has a natural expression in imperative code; both follow directly from the compositional framework.

We adopt the terms \emph{strict} and \emph{lax} by analogy with monoidal functors: a strict monoidal functor preserves the tensor product exactly ($F(A \otimes B) = F(A) \otimes F(B)$), while a lax monoidal functor preserves it only up to a coherence morphism ($F(A \otimes B) \to F(A) \otimes F(B)$). In our setting, the product of island-level Kleisli morphisms is preserved exactly when islands are isolated (strict), and only up to the migration morphism when islands are coupled (lax). The migration rate thus parameterizes a continuous family interpolating between strict and lax composition.

\section{Experimental Setup}

We test the compositional framework across three domains chosen for maximum dissimilarity in genome representation, fitness landscape, and application area. Each domain uses the same four composition strategies, enabling direct comparison of how composition pattern affects population dynamics independent of domain.

\subsection{Domains}

\paragraph{Checkers (real-valued vectors).}
Populations of 30 weight vectors, each encoding 10 Samuel-style board evaluation features (material advantage, king count, back-row defense, center control, advancement, mobility, vulnerability, protection, king centrality, edge presence).
Each weight is a real number in $[-5, 5]$. Fitness is win rate against two baseline opponents (2~games each, alternating colors, 4~games total). Crossover is one-point; mutation is Gaussian perturbation ($\sigma = 0.5$). Tournament selection with size~3.

\paragraph{Maze topology (binary genomes).}
We evolve $6 \times 6$ grid mazes with population size~30. The genome is a 60-bit binary vector encoding wall presence. Fitness is a weighted composite of three BFS-derived metrics: solution path length (longer paths are harder, weight 0.5), dead-end ratio (more dead ends increase difficulty, weight 0.3), and branching factor (more decision points create more interesting navigation, weight 0.2). Crossover is one-point; mutation is bit flip with rate $1/60$ per bit. Tournament selection with size~3.

\paragraph{Symbolic regression (expression trees).}
We evolve populations of 60 expression trees targeting $y = x^2 + x$ from 21 uniformly spaced training points on $[-5, 5]$. The function set is $\{+, -, \times, /\}$ with protected division; the terminal set is $\{x\} \cup \{-5, -4, \ldots, 5\}$. Fitness is negative mean squared error (higher is better). Crossover is subtree exchange; mutation is subtree replacement with maximum depth~4. Tournament selection with size~3. Trees are initialized with ramped half-and-half (depth 2--5).

\paragraph{OneMax (binary, Dichotomy experiment only).}
For the Strict/Lax Dichotomy experiment (Section~\ref{sec:dichotomy}), we use 20-bit OneMax---a standard benchmark where fitness equals the number of 1-bits in a binary string. We use a larger population of~80 (20 per island across 4~islands). The Dichotomy experiment consists of two sub-experiments: a sweep study (40~generations split into two 20-generation phases, varying migration frequency and boundary position) and a multi-seed statistical validation (100~generations across 30~independent random seeds). OneMax provides a smooth, unimodal landscape where all strategies converge to the global optimum, isolating the compositional effect of migration from confounding fitness-landscape effects. We report mean $\pm$ standard deviation across seeds and use the Mann-Whitney $U$ test with Cohen's $d$ effect size to quantify separation between strict and lax compositions.

\subsection{Strategies}

All four strategies compose the same base operators---evaluate, select, crossover, mutate---but differ in \emph{how} they compose them.

\paragraph{Flat generational.}
The standard pipeline: select $\to$ crossover $\to$ mutate $\to$ replace, iterated for 50~generations (20 for checkers/mazes). This is a single Kleisli morphism composed with itself $n$ times. No structural intervention; selection pressure accumulates monotonically.

\paragraph{Hourglass.}
Three-phase sequential composition: \texttt{explore}(15~gen) $\to$ \texttt{converge}(10~gen) $\to$ \texttt{diversify}(25~gen). The explore phase uses high mutation rate ($2\times$ baseline) and weak selection (tournament size~2). The converge phase uses low mutation rate ($0.5\times$ baseline) and strong selection (tournament size~5). The diversify phase returns to explore-phase parameters. This creates a population bottleneck in the middle phase followed by re-expansion---inspired by developmental hourglass models from evolutionary developmental biology.

\paragraph{Island model.}
Four subpopulations arranged in a ring topology, each running the flat generational pipeline independently. Migration occurs every 5~generations: each island sends its best individual to its ring neighbor, replacing the neighbor's worst. The population is split across islands: 8 per island for checkers and mazes (population~30, 4~islands, with two extra individuals assigned to the first two islands), and 15 per island for symbolic regression (population~60). This is a functor from the single-population strategy category to the multi-population strategy category~\cite{whitley1999island, skolicki2005influence}.

\paragraph{Adaptive.}
Begins in \texttt{explore} mode (high mutation, weak selection). A plateau detector monitors fitness improvement over a sliding window of 5~generations. When improvement drops below a threshold ($<1\%$ relative gain), the strategy switches to \texttt{focus} mode (low mutation, strong selection). This is a conditional composition with a runtime predicate---the composition structure depends on the evolutionary trajectory.

\subsection{Metrics}

\paragraph{Genotypic diversity.}
For real-valued genomes (checkers): mean pairwise Euclidean distance, normalized by genome length. For binary genomes (mazes): mean pairwise Hamming distance, normalized to $[0, 1]$. For expression trees (symbolic regression): variance in tree size (node count), which captures structural diversity without requiring a tree edit distance~\cite{burke2004diversity}.

\paragraph{Phenotypic diversity.}
Variance in fitness values across the population. Low phenotypic diversity indicates convergence to a fitness plateau; high phenotypic diversity indicates a spread of solution quality.

\paragraph{Fitness progression.}
Best, mean, and worst fitness per generation, tracked via the logging layer of the evolution monad.

\paragraph{Diversity fingerprint.}
The characteristic shape of the genotypic diversity trajectory over generations. We identify fingerprints qualitatively by their trajectory shape and quantitatively by key features: initial diversity, minimum diversity, final diversity, and the generation at which minimum diversity occurs.

The fingerprint experiments (Sections~4.1--4.2) use single deterministic seeds, with cross-domain consistency across three dissimilar domains providing a form of independent replication (see Section~\ref{sec:limitations}). The Dichotomy experiment (Section~\ref{sec:dichotomy}) uses 30~independent seeds with full statistical testing.


\section{Results}

\subsection{Diversity Fingerprints}

The central empirical finding is that each composition strategy produces a qualitatively distinct diversity trajectory---a \emph{fingerprint}---that is stable across genome types and fitness landscapes. Table~\ref{tab:sr-diversity} shows genotypic diversity over time for all four strategies on symbolic regression, and Figure~\ref{fig:fingerprints} shows the four fingerprint patterns visually.

\begin{table}[ht]
\centering
\caption{Genotypic diversity (tree size variance) over generations for symbolic regression. Bold values mark local extrema characteristic of each fingerprint.}
\label{tab:sr-diversity}
\begin{tabular}{@{}rrrrr@{}}
\toprule
\textbf{Gen} & \textbf{Flat} & \textbf{Hourglass} & \textbf{Island} & \textbf{Adaptive} \\
\midrule
0   & 22.98 & 22.98          & 22.98 & 22.98 \\
5   & 27.71 & \textbf{58.64} & 18.04 & \textbf{52.88} \\
10  & 14.80 & \textbf{44.88} & 11.76 & \textbf{41.81} \\
15  & 11.21 & 15.82          & 6.67  & 4.22 \\
20  & 6.33  & \textbf{3.37}  & 6.24  & 5.43 \\
25  & 6.57  & 4.00           & 7.06  & 3.33 \\
30  & 7.19  & 6.86           & 5.82  & 4.40 \\
50  & 5.71  & 7.07           & 5.83  & 1.92 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[ht]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\columnwidth,
  height=0.65\columnwidth,
  xlabel={Generation},
  ylabel={Genotypic Diversity (tree size variance)},
  xmin=0, xmax=52,
  ymin=0, ymax=65,
  legend pos=north east,
  legend style={font=\small},
  grid=major,
  grid style={gray!30},
  tick label style={font=\small},
  label style={font=\small},
  cycle list name=linestyles,
]

% Flat/Generational — monotonic decline
\addplot[solid, thick, mark=square*, mark size=1.5pt]
  coordinates {(0,22.98) (5,27.71) (10,14.80) (15,11.21) (20,6.33) (25,6.57) (30,7.19) (50,5.71)};

% Hourglass — spike-crash-rebound
\addplot[dashed, thick, mark=triangle*, mark size=2pt]
  coordinates {(0,22.98) (5,58.64) (10,44.88) (15,15.82) (20,3.37) (25,4.00) (30,6.86) (50,7.07)};

% Island — stable maintenance
\addplot[dotted, thick, mark=o, mark size=1.5pt]
  coordinates {(0,22.98) (5,18.04) (10,11.76) (15,6.67) (20,6.24) (25,7.06) (30,5.82) (50,5.83)};

% Adaptive — spike-then-collapse
\addplot[dashdotted, thick, mark=diamond*, mark size=2pt]
  coordinates {(0,22.98) (5,52.88) (10,41.81) (15,4.22) (20,5.43) (25,3.33) (30,4.40) (50,1.92)};

\legend{Flat/Generational, Hourglass, Island, Adaptive}
\end{axis}
\end{tikzpicture}
\caption{Diversity fingerprints for symbolic regression. Each strategy produces a qualitatively distinct trajectory determined by the composition pattern, not the genome type or fitness landscape.}
\label{fig:fingerprints}
\end{figure}

All four strategies reach the same optimum (best fitness $-0.05$, recovering the exact expression $(x \cdot x) + x$), but their diversity dynamics differ dramatically:

\begin{table}[ht]
\centering
\caption{Final-generation comparison across strategies (symbolic regression, 50~generations, population~60).}
\label{tab:sr-final}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Strategy} & \textbf{Best Fitness} & \textbf{Final GenoDiv} & \textbf{Final PhenoDiv} \\
\midrule
Flat      & $-0.05$ & 5.71  & 444.01 \\
Hourglass & $-0.05$ & 7.07  & 277.93 \\
Island    & $-0.05$ & 5.83  & 615.42 \\
Adaptive  & $-0.05$ & 1.92  & 213.42 \\
\bottomrule
\end{tabular}
\end{table}

We identify four distinct fingerprint patterns:

\paragraph{Monotonic decline (flat).}
Diversity drops steadily from 22.98 to 5.71 over 50~generations. No structural intervention counteracts the convergence pressure of tournament selection. The decline is smooth and predictable---each generation removes a fraction of genotypic variation proportional to selection intensity~\cite{blickle1996comparison, crepinsek2013exploration}.

\paragraph{Spike-crash-rebound (hourglass).}
The explore phase drives diversity upward (22.98 $\to$ 58.64 by gen~5). The converge phase crashes it (58.64 $\to$ 3.37 by gen~20---a 94\% reduction). The diversify phase partially restores it (3.37 $\to$ 7.07 by gen~50). The three-phase structure of the composition is directly visible in the diversity trajectory. The crash point at gen~15--20 corresponds exactly to the converge-phase boundary.

\paragraph{Stable maintenance (island).}
After an initial adjustment (22.98 $\to$ 18.04 by gen~5), diversity stabilizes in the 5.8--7.1 range for the remaining 45~generations. Migration every 5~generations injects just enough genetic material to counterbalance selection-driven convergence within each subpopulation. The island model acts as a diversity thermostat~\cite{whitley1999island}.

\paragraph{Spike-then-collapse (adaptive).}
The explore phase spikes diversity (22.98 $\to$ 52.88 by gen~5), similar to the hourglass explore phase. But when the plateau detector triggers the switch to focus mode, diversity collapses irreversibly to 1.92---the lowest final diversity of any strategy. Unlike the hourglass, there is no recovery phase. The adaptive strategy trades long-term diversity for rapid convergence once a promising region is found.

The fingerprint is determined by the \emph{composition pattern}, not by the operators.
All four strategies use identical selection, crossover, and mutation operators.
Only how these operators are composed---sequentially, in phases, across islands, or conditionally---differs.
Yet this compositional difference produces a $3.7\times$ spread in final genotypic diversity (1.92 to~7.07) and a $2.9\times$ spread in final phenotypic diversity (213.42 to~615.42).


\subsection{Cross-Domain Stability}

To test whether fingerprints are composition-invariant rather than domain-specific, we apply flat, hourglass, and island strategies to checkers and mazes (Table~\ref{tab:cross-domain}).

\begin{table}[ht]
\centering
\caption{Cross-domain genotypic diversity over generations. C = Checkers (Euclidean, normalized), M = Mazes (Hamming, normalized). Diversity is measured on domain-appropriate scales.}
\label{tab:cross-domain}
\begin{tabular}{@{}rrrrrrr@{}}
\toprule
\textbf{Gen} & \textbf{C:Flat} & \textbf{C:Hrgl} & \textbf{C:Isld} & \textbf{M:Flat} & \textbf{M:Hrgl} & \textbf{M:Isld} \\
\midrule
0  & 0.49 & 0.49 & 0.49 & 0.47 & 0.47 & 0.47 \\
5  & 0.34 & 0.32 & 0.23 & 0.21 & 0.51 & 0.37 \\
10 & 0.33 & 0.33 & 0.47 & 0.26 & 0.29 & 0.31 \\
15 & 0.31 & 0.32 & 0.44 & 0.26 & 0.46 & 0.28 \\
20 & 0.30 & 0.30 & 0.34 & 0.27 & 0.46 & 0.32 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Final comparison across two domains and three strategies (20~generations, population~30).}
\label{tab:final-comparison}
\begin{tabular}{@{}llrrr@{}}
\toprule
\textbf{Domain} & \textbf{Strategy} & \textbf{Best Fit} & \textbf{GenoDiv} & \textbf{PhenoDiv} \\
\midrule
Checkers & Flat      & 1.0  & 0.30 & 0.0 \\
Checkers & Hourglass & 1.0  & 0.30 & 0.0 \\
Checkers & Island    & 1.0  & 0.34 & 0.0 \\
Maze     & Flat      & 0.70 & 0.27 & 40.36 \\
Maze     & Hourglass & 0.68 & 0.46 & 19.96 \\
Maze     & Island    & 0.73 & 0.32 & 40.14 \\
\bottomrule
\end{tabular}
\end{table}

The qualitative ordering of final genotypic diversity---flat $<$ island $<$ hourglass---holds for mazes ($0.27 < 0.32 < 0.46$) and symbolic regression ($5.71 < 5.83 < 7.07$). Checkers deviates: flat and hourglass tie at $0.30$, with island slightly higher at $0.34$. This deviation is expected---the checkers fitness landscape saturates rapidly (all strategies achieve perfect 1.0 win rate), compressing diversity differences. On non-saturating landscapes, the island model consistently maintains intermediate diversity while the hourglass produces the highest final diversity but sacrifices fitness (0.68 vs.\ 0.73 island in mazes), consistent with the diversity-quality tradeoff predicted by the spike-crash-rebound fingerprint.

\paragraph{Multi-domain topology ordering.}
A topology sweep across five domains---OneMax (binary), maze navigation (binary), graph coloring (integer), knapsack (binary), and No Thanks! (co-evolutionary, tournament-based relative fitness)---reveals that composition structure determines diversity dynamics with striking universality. Island-model GAs were run on five migration topologies (none, ring, star, random, fully connected) with 30 independent seeds per condition. All five domains produce the identical mean-fitness ordering:
\begin{center}
\small
none $<$ ring $<$ star $<$ random $<$ fully\_connected
\end{center}
\noindent with pairwise Spearman rank correlation $\rho = 1.0$ across all $\binom{5}{2} = 10$ domain pairs. A two-way ANOVA finds that topology explains $23.9\times$ more variance in mean fitness than domain ($F_{\mathrm{topo}} = 47.8$, $p < 10^{-6}$; $F_{\mathrm{domain}} = 0.13$, $p = 0.945$). The domain factor is non-significant: it contributes essentially no explanatory power beyond topology. The No Thanks! result is especially significant: as a co-evolutionary domain with tournament-based relative fitness, there is no fixed fitness landscape---yet the canonical topology ordering persists ($\rho = 1.0$, phase transition 53.9\%), providing the strongest domain-independence evidence to date. This ordering follows the algebraic connectivity $\lambda_2$ of the migration graph~\cite{sanz2026algebraic}, confirming a structural prediction from coupled oscillator theory. A sixth domain (sorting networks) violates the ordering, providing a falsifiable scope condition: the invariant holds for domains where the fitness landscape admits sufficient selective gradient, and breaks where domain-specific constraints dominate migration dynamics.

A separate $n{=}7$ island experiment on mazes confirms the spectral theorem prediction at the boundary where the ring--star ordering changes sign: at $n{\geq}7$ islands, $\lambda_2(\text{ring}) < \lambda_2(\text{star})$ and star outperforms ring ($p = 6.6{\times}10^{-5}$), validating the algebraic connectivity prediction independently of the $n{=}5$ sweep.

Furthermore, early convergence analysis across domains reveals a two-phase process: coupling onset occurs universally at generations 5--7 (independent of domain), followed by gradual differentiation, with the topology ordering fully established by generation~30. This onset timing matches the spectral gap prediction from coupled oscillator theory~\cite{sanz2026algebraic}.

The key result is that \emph{composition structure is the invariant}. Across five domains spanning binary, integer, real-valued, and co-evolutionary genomes, the same topological composition produces the same diversity and fitness ordering. The fingerprint follows the functor, not the object.


\subsection{The Strict/Lax Dichotomy}
\label{sec:dichotomy}

The most practically significant result concerns the island model's behavior under sequential composition. In operational terms: \emph{if you checkpoint an island-model GA and restart it, do you get the same result as running it continuously?}

The answer is: \textbf{only if migration rate is zero.}

We test this on a 4-island ring topology evolving 20-bit OneMax (population~80, 20 per island, 40~generations split into two 20-generation phases). We compare $I(S_1 ; S_2)$ (continuous 40-generation run) against $I(S_1) ; I(S_2)$ (two independent 20-generation runs composed sequentially, resetting the migration schedule at the boundary).

\paragraph{Migration frequency sweep.}
Table~\ref{tab:laxity} varies migration frequency while holding all other parameters constant.

\begin{table}[ht]
\centering
\caption{Island functor laxity vs.\ migration frequency. Freq = generations between migration events. Sched Diff = number of migration events displaced by the composition boundary. Pop Divergence = normalized distance between continuous and composed final populations.}
\label{tab:laxity}
\small
\begin{tabular}{@{}rrrrl@{}}
\toprule
\textbf{Freq} & \textbf{Sched$\Delta$} & \textbf{Pop Div} & \textbf{Ham Div} & \textbf{Fit$\Delta$} \\
\midrule
2  & 1 & 0.744          & 0.110 & 0.0 \\
5  & 1 & 0.812          & 0.102 & 0.0 \\
10 & 1 & 0.756          & 0.096 & 0.0 \\
20 & 1 & 0.819          & 0.129 & 0.0 \\
\textbf{40} & \textbf{0} & \textbf{0.000} & \textbf{0.000} & \textbf{0.0} \\
\bottomrule
\end{tabular}
\end{table}

The transition is binary. At freq=40 (no migration occurs within the 40-generation window), the functor is \textbf{strict}: continuous and composed runs produce identical populations (divergence = 0.000). At every nonzero migration frequency, the functor is \textbf{uniformly lax} with population divergence ${\sim}0.78 \pm 0.04$, independent of how often migration occurs. Migrating every 2~generations produces the same divergence magnitude as migrating every 20. Fitness is unaffected in all conditions (all reach OneMax optimum), confirming that the divergence is genotypic, not phenotypic.

\paragraph{Boundary position sweep.}
Table~\ref{tab:boundary} varies where the composition boundary falls relative to the migration schedule (freq=5, boundary positions from gen~15 to gen~25).

\begin{table}[ht]
\centering
\caption{Divergence vs.\ composition boundary position (migration freq=5). ``Hits migration'' indicates whether the boundary coincides with a scheduled migration event.}
\label{tab:boundary}
\small
\begin{tabular}{@{}rlrr@{}}
\toprule
\textbf{Boundary} & \textbf{Hits Migr?} & \textbf{Pop Div} & \textbf{Ham Div} \\
\midrule
15 & YES & 0.794 & 0.105 \\
17 & no  & 0.725 & 0.098 \\
20 & YES & 0.812 & 0.102 \\
23 & no  & 0.744 & 0.108 \\
25 & YES & 0.800 & 0.102 \\
\bottomrule
\end{tabular}
\end{table}

Divergence is \textbf{uniform} (${\sim}0.78 \pm 0.04$) regardless of whether the boundary coincides with a migration event. This falsifies the naive hypothesis that schedule disruption matters only when the boundary directly displaces a migration event. The resetting of the migration counter at the boundary is sufficient; the perturbation is amplified to a characteristic magnitude $D^*$ by the chaotic dynamics of the interacting subpopulations.

\paragraph{Multi-seed statistical validation.}
To confirm the Dichotomy with rigorous statistics, we run 30~independent seeds comparing strict (no migration) and lax (migration every 5~generations) island compositions on 20-bit OneMax over 100~generations (population~80, 4~islands of~20). Table~\ref{tab:multiseed} reports mean $\pm$ standard deviation of normalized Hamming diversity, with Mann-Whitney $U$ test $p$-values and Cohen's $d$ effect sizes.

\begin{table}[ht]
\centering
\caption{Strict vs.\ lax composition diversity (mean $\pm$ std) over 30~independent seeds (20-bit OneMax, 100~gen). Cohen's $d > 0.8$ is ``large''; $d > 2.0$ indicates near-complete separation.}
\label{tab:multiseed}
\small
\begin{tabular}{@{}lrrrl@{}}
\toprule
\textbf{Gen} & \textbf{Strict} & \textbf{Lax} & \textbf{$d$} & \textbf{$p$} \\
\midrule
0   & $.500 \pm .001$ & $.500 \pm .001$ & 0.00 & $1.00$ \\
10  & $.243 \pm .034$ & $.228 \pm .041$ & 0.41 & $1.1{\times}10^{-1}$ \\
25  & $.117 \pm .014$ & $.099 \pm .019$ & 1.05 & $4.0{\times}10^{-4}$ \\
50  & $.048 \pm .008$ & $.028 \pm .008$ & 2.55 & $4.2{\times}10^{-10}$ \\
75  & $.037 \pm .004$ & $.017 \pm .004$ & 5.36 & $3.0{\times}10^{-11}$ \\
100 & $.035 \pm .005$ & $.016 \pm .004$ & 4.34 & $3.7{\times}10^{-11}$ \\
\bottomrule
\end{tabular}
\end{table}

The results are unambiguous. By generation~50, strict composition maintains $1.72\times$ the diversity of lax composition ($0.048 \pm 0.008$ vs.\ $0.028 \pm 0.008$), with Cohen's $d = 2.55$ indicating near-complete separation of the 30-seed distributions. By generation~100, the effect size reaches $d = 4.34$ ($p = 3.7 \times 10^{-11}$): the strict and lax distributions do not overlap. Crucially, \textbf{both compositions converge to the same fitness optimum}---strict and lax mean best fitness both reach 100.0 by generation~57. The divergence is purely in diversity dynamics, exactly as the Dichotomy predicts: composition pattern determines the diversity fingerprint but not the fitness outcome.

The statistical figure (mean trajectories with 95\% confidence bands across 30~seeds) confirms that the separation between strict and lax diversity trajectories is not a single-seed artifact but a robust, large-effect phenomenon. The confidence bands for the two compositions cease to overlap by approximately generation~20 and remain fully separated thereafter.

\begin{conjecture}[Strict/Lax Dichotomy]
\label{thm:dichotomy}
Let $I(\mu, \mathrm{freq}, n, \mathrm{topo})$ be the island strategy functor parameterized by migration count $\mu$, migration frequency, island count $n$, and migration topology. For any strategies $S_1$, $S_2$:
\begin{enumerate}[nosep]
  \item If $\mu = 0$ or $\mathrm{freq} > |S_1| + |S_2|$ (no migration events occur), then $I(S_1 ; S_2) = I(S_1) ; I(S_2)$---the functor is \textbf{strict}.
  \item Otherwise, $I(S_1 ; S_2) \cong I(S_1) ; I(S_2)$ with laxity magnitude $D^*$ that is independent of boundary position and migration frequency, determined asymptotically by the spectral gap of the migration topology's Markov chain. The multi-seed experiment confirms this with Cohen's $d$ up to 5.36 across 30~independent runs (Table~\ref{tab:multiseed}).
\end{enumerate}
\end{conjecture}

\paragraph{Practical implication.}
This result has direct consequences for distributed GA implementations. Any system that checkpoints and restarts island-model runs, composes partial runs from different compute nodes, or chains island-model phases in a multi-stage optimization pipeline must account for the laxity. The composed result will differ from continuous execution by a characteristic magnitude $D^*$---not by a small perturbation, but by a population-level divergence of ${\sim}0.78$ in normalized terms. For applications where genotypic reproducibility matters (e.g., evolved safety-critical controllers), this is a dealbreaker for naive checkpointing. For applications where only fitness matters, the laxity is benign: fitness divergence is 0.0 across all conditions tested. The multi-seed experiment (Table~\ref{tab:multiseed}) confirms this at scale: both strict and lax compositions reach the OneMax optimum, yet their diversity trajectories diverge with effect sizes exceeding $d = 4.0$ from generation~50 onward. This pattern echoes findings in LLM-based multi-agent systems, where independent (lax) composition amplifies errors $17.2\times$ versus $4.4\times$ for centralized (strict) composition~\cite{kim2025scaling}. Han et al.~\cite{han2026conformity} observe analogous dynamics in LLM social conformity: hub-topology agents converge fast but are fragile to perturbation (strict), while distributed agents converge slowly but are robust (lax).

\subsection{Evolved Artifacts}

To ground the fingerprint results, we briefly examine artifact quality.
In checkers, all strategies saturate to perfect fitness (1.0 win rate), so fingerprint differences do not translate into quality differences.
In mazes, the island strategy produces the highest-fitness mazes (0.73 vs.\ 0.70 flat, 0.68 hourglass)---its diversity maintenance fingerprint preserves structural variants that flat composition eliminates.
In symbolic regression, all strategies recover the exact target $(x \cdot x) + x$, but final phenotypic diversity ranges from 213.42 (adaptive) to 615.42 (island), a $2.9\times$ spread.
The composition pattern determines the population's diversity profile even when fitness outcomes are identical.

\section{Related Work}

\paragraph{GA Theory and Composition.}
Classical GA theory analyzes operators individually. Holland's schema theorem~\cite{holland1975adaptation} and Goldberg's building block hypothesis~\cite{goldberg1989genetic, goldberg2002design} characterize how selection and crossover process schemata, but treat these operators in isolation---the theory says nothing about what happens when you compose them differently. The royal road experiments~\cite{mitchell1992royal} showed that the relationship between operators and fitness landscapes is subtler than the building block hypothesis predicts, but still analyzed crossover as a standalone mechanism. Island model theory~\cite{whitley1999island, skolicki2005influence} studies migration rates and topology, yet treats migration as a parameter rather than a composition functor whose interaction with sequential composition can be formally characterized. Population diversity management~\cite{burke2004diversity, crepinsek2013exploration} provides metrics for measuring diversity and frameworks for understanding the exploration-exploitation tradeoff, but does not connect diversity dynamics to the compositional structure of the algorithm. Our work shows that \emph{composition pattern determines diversity trajectory}---a claim that requires making composition explicit, which none of these frameworks do.

\paragraph{Categorical Methods in Optimization.}
Category theory has recently been applied to several optimization paradigms.
Bakirtzis et al.~\cite{bakirtzis2024compositional} give a compositional account of reinforcement learning via categorical MDPs (JMLR 2025), proving safety guarantees that compose across subsystems---the closest precedent for our compositionality-first approach.
Hedges and Sakamoto~\cite{hedges2024reinforcement} model reinforcement learning as parametrised optics (EPTCS 2025).
Gavranovi\'{c} et al.~\cite{gavranovic2024categorical} formalize neural network architectures as monads in the Para construction (ICML 2024).
Hanks et al.~\cite{hanks2024gradient} cast gradient descent as a hypergraph functor.
Miranda~\cite{miranda2024abstract} develops abstract Kleisli structures that generalize the construction we instantiate here.
Ghani et al.~\cite{ghani2025approximate} develop approximate game theory using categorical methods; their treatment of strict vs.\ approximate preservation parallels our Strict/Lax Dichotomy structurally.
Li et al.~\cite{li2023emergence} develop a sheaf-theoretic framework for emergence via derived functors, where cohomological obstructions measure failure of local-to-global extension---the same phenomenon our laxator captures from a different mathematical tradition.
Our contribution fills the remaining gap: \emph{evolutionary computation} has had no categorical treatment of operator composition until now. Taken together, these results span the optimization landscape---neural networks, reinforcement learning, gradient methods, and now GAs---all within categorical frameworks.

\paragraph{Categorical Approaches to Evolutionary Biology.}
Warrell et al.~\cite{warrell2024kantorovich} apply the Kantorovich monad to model multilevel selection in evolutionary biology, using Wasserstein distances on probability distributions to capture population hierarchy.
Their work is complementary to ours: they model the \emph{structure} of populations (how subpopulations nest), while we model the \emph{composition} of operators (how selection, crossover, and mutation chain together as Kleisli arrows).
Their framework does not represent operators as morphisms and has no analog of our diversity fingerprints or the Strict/Lax Dichotomy.
Together, the two approaches suggest that categorical methods can capture both the static hierarchy and the dynamic composition of evolutionary systems.

\paragraph{LLM-Driven Evolution and Scaling.}
Recent work uses large language models as evolutionary operators.
AlphaEvolve~\cite{alphaevolve2026marl} drops crossover entirely when the LLM mutation monad is sufficiently expressive---a composition choice with predictable fingerprint consequences in our framework.
CodeEvolve~\cite{assumpcao2025codeevolve} uses three modular LLM operators on an island-based GA: depth exploitation (strict---preserves lineage), inspiration-based crossover (intermediate---synthesizes features across parents), and meta-prompting exploration (lax---deliberately excludes ancestor chains). The full composition outperforms any individual operator and their sum on most benchmarks, exhibiting genuine compositional synergy rather than mere aggregation.
Kim et al.~\cite{kim2025scaling} find error amplification of $17.2\times$ for independent (lax) vs.\ $4.4\times$ for centralized (strict) multi-agent compositions---a numerical shadow of our Dichotomy.
Van Stein et al.~\cite{vanstein2025code} show that different evolutionary strategies produce distinct search trajectory network signatures determined by operator composition, not problem instance---our thesis in graph-theoretic language.

\paragraph{Topology Optimization in Multi-Agent Systems.}
Yang et al.~\cite{yang2025topology} argue that topological structure learning is an underresearched priority for LLM-based multi-agent systems, documenting up to 10\% performance variation across fixed topologies. Several empirical approaches address this: Zhou et al.~\cite{zhou2025mass} decompose multi-agent design into prompt and topology optimization, finding 10--15\% improvement from topology alone; Cang et al.~\cite{cang2026graphgrpo} use reinforcement learning to discover optimal communication graphs; and Li et al.~\cite{li2026ofamas} train a generative model that produces task-specific topologies. These learn \emph{which} topologies work through search; our framework explains \emph{why} via the spectral bridge and Strict/Lax Dichotomy.

\paragraph{Monads in Programming Language Theory.}
Moggi~\cite{moggi1991notions} introduced monads as a unifying framework for computational effects---randomness, state, I/O---in programming language semantics. This idea, now standard in functional programming, provides the foundation for our evolution monad. We instantiate Moggi's framework for evolutionary computation: the monad carries randomness, configuration, and logging, and Kleisli composition gives us the compositionality that imperative GA implementations lack. To our knowledge, this is the first application of Kleisli categories to genetic algorithm analysis.

\section{Discussion and Conclusion}

\subsection{Practical Implications}

The compositional framework yields four concrete recommendations for GA practitioners.

\paragraph{Fingerprint-guided strategy design.}
If you observe monotonic diversity decline during a GA run, your composition is generational---selection pressure is accumulating without structural counterbalance. Our results show that switching to island composition maintains diversity in the 5.8--7.1 range (Table~\ref{tab:sr-diversity}) while achieving equal or higher fitness, and hourglass composition produces even higher final diversity (7.07 vs.\ 5.71) at the cost of a mid-run bottleneck. The diversity fingerprint is a diagnostic: it tells you which composition pattern your algorithm is executing, and therefore which interventions will change its trajectory. Monitoring genotypic diversity over time is cheap; the fingerprint shape immediately identifies the composition class.

\paragraph{Checkpointing and the Dichotomy.}
The Strict/Lax Dichotomy has a direct operational consequence: \emph{checkpointing an island-model GA and resuming it is not equivalent to running it continuously}, unless migration rate is zero. Any nonzero migration produces population divergence of ${\sim}0.78$ in normalized terms (Table~\ref{tab:laxity}), independent of migration frequency or boundary position. The 30-seed experiment (Table~\ref{tab:multiseed}) confirms this is not a single-seed artifact: strict and lax compositions maintain statistically distinct diversity levels with Cohen's $d > 4.0$ ($p < 10^{-10}$) from generation~50 onward. Practitioners who checkpoint island-model runs for fault tolerance, distribute phases across compute nodes, or chain island-model stages in multi-phase optimization pipelines are introducing systematic bias. For applications where genotypic reproducibility matters (evolved controllers, safety-critical systems), this bias must be accounted for. For applications where only fitness matters, the laxity is benign: fitness divergence is 0.0 across all 30~seeds and all conditions tested.

\paragraph{Design algebra.}
The three-level composition tower (operators $\to$ pipelines $\to$ strategies) provides a principled alternative to ad-hoc strategy design. Instead of writing a custom evolutionary loop and hoping it works, compose pipelines from typed operators using Kleisli composition, then compose strategies from pipelines using the same mechanism. The monad laws guarantee associativity---re-parenthesizing a composition does not change behavior---and the type system catches composition errors at compile time. This is the difference between ``I wrote a GA loop that seems to work'' and ``I composed a pipeline whose properties follow from its structure.''

\paragraph{Cross-domain portability.}
Since fingerprints are composition-determined rather than domain-determined (Section~4.2), insights transfer across domains. If island composition maintains diversity on a cheap problem (e.g., OneMax or small mazes), it will maintain diversity on an expensive problem with the same composition structure (e.g., large-scale combinatorial optimization). Practitioners can tune their \emph{composition structure} on inexpensive benchmarks and apply it to expensive target problems with confidence that the qualitative diversity dynamics will transfer. The composition is the invariant; the domain is a parameter.

\paragraph{Predictive power.}
The framework is not merely post-hoc: it generates testable predictions for untested configurations. Any migration topology with algebraic connectivity between ring and fully connected should exhibit intermediate diversity dynamics, regardless of the fitness landscape. This prediction aligns with three independent confirmations from adjacent fields: Sanz~\cite{sanz2026algebraic} shows that $\lambda_2$ universally governs synchronization onset in coupled oscillators; Wu and Brandes~\cite{wu2025threshold} prove that threshold graphs (of which the complete graph is a special case) are globally synchronizing---our ``fully connected = maximally strict'' result in graph-theoretic language; and Ostilli~\cite{ostilli2025topological} demonstrates that local network structure determines synchronization robustness, paralleling our finding that topology determines diversity dynamics. Brewster et al.~\cite{brewster2025evolutionary} provide a fourth confirmation from evolutionary graph theory: fixation times scale as $N^2$ on complete graphs versus $N^3$ on rings, exactly the strict/lax ordering our framework predicts. More broadly, any system whose operator composition can be typed as Kleisli arrows over a population monad inherits the fingerprint and dichotomy predictions without domain-specific re-derivation.

\subsection{Limitations and Future Work}
\label{sec:limitations}

\paragraph{Single-seed fingerprint experiments.}
The fingerprint results (Sections~4.1--4.2) use single random seeds per configuration. While the cross-domain consistency across three dissimilar domains (real-valued, binary, tree-structured) provides a form of independent replication, proper statistical validation requires multi-seed runs. The Dichotomy experiment (Section~\ref{sec:dichotomy}) addresses this with 30~independent seeds and full statistical testing (Mann-Whitney $U$, Cohen's $d$), confirming that composition-determined diversity differences are robust ($d > 4.0$, $p < 10^{-10}$). Extending multi-seed validation to the fingerprint experiments across all three domains remains future work.

\paragraph{Domain coverage.}
The topology sweep (Section~4.2) extends domain coverage to seven problems (OneMax, maze, graph coloring, knapsack, No Thanks!, checkers, sorting networks), with five showing identical topology ordering. Extending to permutation-based domains (TSP, scheduling), variable-length genomes (NEAT-style neuroevolution~\cite{eiben2015introduction}), and high-dimensional continuous optimization would further test universality. The sorting network exception suggests that domain-specific constraints can override compositional effects---characterizing this boundary precisely remains future work.

\paragraph{Formal proofs.}
The Strict/Lax Dichotomy (Conjecture~\ref{thm:dichotomy}) is stated with formal precision and supported by experimental evidence across migration frequencies and boundary positions. Full categorical proofs, including the connection between laxity magnitude $D^*$ and the spectral gap of the migration topology's Markov chain, are developed in an extended version of this work. The present paper focuses on the empirical phenomenon and its practical consequences.

\paragraph{The co-Kleisli direction.}
Our framework models what operators \emph{do}: each operator takes a population and produces a new population within an effect context. The dual perspective---what operators \emph{need} from their environment---would use co-Kleisli arrows of a comonad, where the comonadic context represents the population structure that each operator depends on. This co-Kleisli formulation could model selection as a context-dependent extraction (the comonad provides the population; selection extracts from it) and would give a bidirectional account of evolutionary computation. We leave this direction for future work.

\paragraph{Mainstream integration.}
Our implementation uses Haskell, where monads and Kleisli composition are first-class language features.
Porting to mainstream EC frameworks (DEAP, ECJ, jMetal) would require categorical libraries or disciplined use of interfaces and type classes.
The compositional structure itself is language-agnostic; accessibility is an engineering challenge, not a theoretical one.

\subsection{Conclusion}

Composition determines behavior. We have demonstrated this across seven domains with three results: diversity fingerprints are stable signatures of composition pattern, not of individual operators or genome types---with topology explaining $23.9\times$ more variance than domain across five independent problems; the island functor exhibits a sharp strict/lax dichotomy at zero migration---confirmed across 30~independent seeds with Cohen's $d > 4.0$ ($p < 10^{-10}$)---with practical consequences for checkpointing and distributed execution; and three levels of composition---operators, pipelines, strategies---share the same Kleisli algebraic structure, giving a recursive design algebra for genetic algorithms.

While Warrell et al.~\cite{warrell2024kantorovich} apply categorical methods to population hierarchy in evolutionary biology, our work is the first to treat GA \emph{operator composition} categorically. By formalizing operators as Kleisli arrows and strategies as higher-order operations on these arrows, we bring evolutionary algorithms into the same mathematical framework that now encompasses neural networks~\cite{gavranovic2024categorical}, reinforcement learning~\cite{bakirtzis2024compositional, hedges2024reinforcement}, and gradient methods~\cite{hanks2024gradient}. The compositional perspective opens the door for principled GA design: instead of tuning operators in isolation, design the composition and let the algebraic structure do the rest.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
