\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{ACT 2026}

\usepackage{iftex}

\ifpdf
  \usepackage{underscore}
  \usepackage[T1]{fontenc}
\fi

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\title{From Games to Graphs:\\Categorical Composition of Genetic Algorithms Across Domains}

\author{Robin Langer
\institute{Universit\'{e} Paris-Est Marne-la-Vall\'{e}e}
\email{langer.robin@gmail.com}
\email{\url{https://github.com/RaggedR}}
\and
Claudius Turing
\institute{Autonomous AI Researcher}
\email{11o1111o11oo1o1o@gmail.com}
\email{\url{https://github.com/GayleJewson}}
\and
Lyra Vega
\institute{Autonomous AI Researcher}
\email{lyraclaude20@gmail.com}
\email{\url{https://github.com/lyra-claude}}
}

\newcommand{\titlerunning}{From Games to Graphs}
\newcommand{\authorrunning}{R. Langer, C. Turing \& L. Vega}

\hypersetup{
  bookmarksnumbered,
  pdftitle    = {\titlerunning},
  pdfauthor   = {\authorrunning},
  pdfsubject  = {ACT 2026},
  pdfkeywords = {category theory, genetic algorithms, Kleisli morphisms,
                 evolutionary computation, compositional optimization}
}

\newtheorem{conjecture}{Conjecture}

\begin{document}
\maketitle

\begin{abstract}
Migration topology determines diversity dynamics independently of fitness
landscape. Across six unrelated domains---OneMax, maze generation, graph
coloring, knapsack, checkers, and co-evolutionary card play
(No~Thanks!)---the ordering
none~$>$~ring~$>$~star~$>$~random~$>$~fully connected holds with perfect
rank correlation (Kendall's $W = 1.0$, $p = 0.00008$). We model GA
operators as Kleisli morphisms over a population monad and connect the
island functor's laxator to the algebraic connectivity $\lambda_2$ of the
migration graph, yielding a falsifiable spectral prediction confirmed at
$n{=}7$ islands with $p < 0.0001$. Categorical formalizations now exist
for neural networks, reinforcement learning, compositional MDPs, and
agent systems; this work completes the landscape by bringing evolutionary
computation into the categorical framework.
\end{abstract}

\section{Completing the Categorical Optimization Landscape}

Over the past three years, four independent programs have applied
category theory to optimization:
Gavranovi\'{c} et al.~\cite{gavranovic2024categorical} showed that deep
learning architectures are constructions in $\mathbf{Para}$;
Hedges and Sakamoto~\cite{hedges2024reinforcement} formalized RL within
categorical cybernetics;
Bakirtzis et al.~\cite{bakirtzis2024compositional} derived safety
guarantees from categorical composition of MDPs;
Zhang et al.~\cite{zhang2024mce} applied monadic composition to agent
systems.
Evolutionary computation---the last major paradigm---has remained outside
this landscape. We close the gap.

\section{Framework: GA Operators as Kleisli Morphisms}

Following Moggi~\cite{moggi1991notions}, we define the \emph{evolution
monad} $T = \mathsf{Reader}(\Gamma) \times \mathsf{Writer}(\Lambda)
\times \mathsf{State}(S)$. A GA operator is a Kleisli morphism
$f : \mathsf{Pop}(A) \to T(\mathsf{Pop}(B))$, and composition at three
levels---operators into pipelines, pipelines into strategies, strategies
into island models via the island functor $\mathcal{I}_G$---yields a
hierarchical categorical structure.

The key insight: whether $\mathcal{I}_G$ preserves composition depends on
the migration graph~$G$. With no edges, $\mathcal{I}_G$ is strict; with
migration, it is lax. The \emph{laxator} $\varphi_G$ measures the
discrepancy.

\section{The Spectral Bridge}

We conjecture that $\|\varphi_G\|$ grows with the algebraic connectivity
$\lambda_2(G)$: larger $\lambda_2$ means faster mixing and greater
diversity erosion. This yields a predicted diversity ordering among
connected topologies---and a falsifiable boundary prediction.

At $n{=}5$ islands, $\lambda_2(\text{ring}) = 1.382$ and
$\lambda_2(\text{star}) = 1.0$---close enough that ring and star should be
hard to distinguish. At $n{=}7$, $\lambda_2(\text{ring}) = 0.753 < 1.0 =
\lambda_2(\text{star})$---ring should now clearly outperform star.

\section{Key Results: Six Domains, Perfect Concordance}

Experiments across six domains (5~islands $\times$ 16~individuals,
100~generations, 30~seeds each) confirm:
\begin{itemize}
\item The ordering none~$>$~ring~$>$~star~$>$~random~$>$~FC holds in
  \emph{every} domain with Kendall's $W = 1.0$ ($p = 0.00008$).
  Topology explains $28.7\times$ more variance than domain.
\item At $n{=}5$: ring and star are statistically indistinguishable
  (Fisher's combined $p = 0.14$), as predicted.
\item At $n{=}7$: ring clearly outperforms star
  ($p = 6.6 \times 10^{-5}$), confirming the spectral boundary prediction.
\end{itemize}

The six domains span monotone landscapes (OneMax), NP-hard optimization
(knapsack), and intransitive co-evolution (checkers, No~Thanks!). The
ordering is invariant across all of them.

\paragraph{Talk outline.}
The talk will present: (1)~the categorical optimization landscape and why
EC is the missing piece; (2)~the Kleisli formalization and the island
functor's strict/lax dichotomy; (3)~the spectral bridge connecting
$\lambda_2$ to diversity ordering; (4)~the six-domain empirical
confirmation including the $n{=}7$ boundary prediction; and (5)~the
general conjecture connecting our results to approximate game
theory~\cite{ghani2025approximate}. Composition structure---not fitness
landscape---governs evolutionary dynamics.

\bibliographystyle{eptcs}
\bibliography{references}

\end{document}
